% chapter10.tex -- Finding the Safe Root in O(n) Time (Two-Pointer Algorithm)
% Based on the author's UGThesis.docx.

\chapter{Finding the Safe Root in \(O(n)\) Time (Two-Pointer Algorithm)}
\label{ch:safe_root_algorithm}

In this chapter we present an efficient algorithm to find a safe root vertex for a face in linear time. 
A brute-force approach would test each vertex as a candidate and check all its potential chords against the global set \(S\), resulting in \(O(n^2)\) time per face, which would violate the amortised constant-time guarantee. 
Using planarity, we design a two-pointer algorithm that scans the cycle once and identifies a safe root in \(O(n)\) time.

\section{Problem Restatement}
\label{sec:safe_root_problem}

Let \(F\) be a face (a simple cycle) with vertices \(v_1, v_2, \dots, v_n\) in counterclockwise order. 
Let \(S\) be the global chord set containing all chords from previously triangulated faces that lie outside \(F\) (and also the permanent edges of the graph). 
We need to find a vertex \(v_r\) such that for every non-neighbouring vertex \(v_j\) of \(v_r\) on the cycle, the chord \((v_r, v_j)\) is not already in \(S\). 
Such a vertex is called a \textbf{safe root} (Definition~\ref{def:safe_root}).

The existence of at least two safe roots was proved in Theorem~\ref{thm:safe_root_exists}. 
Now we focus on the algorithmic task of finding one efficiently.

\section{Brute Force and Its Limitations}
\label{sec:brute_force}

A straightforward method would iterate over each vertex \(v_i\) as a candidate root and, for each such candidate, check all chords \((v_i, v_j)\) with \(j \neq i-1, i, i+1\) (mod \(n\)) for membership in \(S\). 
If none of these chords is found in \(S\), then \(v_i\) is safe.

There are \(n\) candidates, and each candidate requires checking up to \(n-3\) chords. 
Thus the brute-force time is \(O(n^2)\). 
Since the total number of triangulations for a face can be as low as \(\Omega(n)\) (Chapter~\ref{ch:lower_bound}), an \(O(n^2)\) building cost would dominate the total work, breaking the \(O(1)\) amortised guarantee. 
Therefore we need a linear-time method.

\section{Two-Pointer Algorithm Based on Planarity}
\label{sec:two_pointer}

The key idea is to use planarity to prune the search. 
If we find a chord \((v_i, v_j)\) that belongs to \(S\) (i.e., an external chord already present), then by planarity, no vertex strictly between \(v_i\) and \(v_j\) along the cycle can have a chord to a vertex strictly outside that interval without crossing \((v_i, v_j)\). 
This observation allows us to move pointers inward.

We maintain two pointers: 
\begin{itemize}
    \item \(left\) (initially \(1\)) – the current candidate root.
    \item \(right\) (initially \(n-1\)) – the vertex we test chords against.
\end{itemize}
We test whether the chord \((v_{left}, v_{right})\) exists in \(S\). 
If it does, then \(v_{left}\) cannot be a safe root (because this chord would conflict), so we increment \(left\) (move to the next candidate). 
If the chord does not exist in \(S\), we decrement \(right\) (try a vertex further away). 
The process continues until \(left\) and \(right\) become neighbours (i.e., \(right = left+1\)), at which point \(v_{left}\) is guaranteed to be a safe root.

\begin{algorithm}[htbp]
    \caption{FindSafeRoot (Two-pointer method)}
    \label{alg:find_safe_root}
    \begin{algorithmic}[1]
        \Procedure{FindSafeRoot}{$F, S$}
            \State Let the vertices be $v_1, v_2, \dots, v_n$ in counterclockwise order.
            \State $left \gets 1$
            \State $right \gets n-1$  \Comment{$v_n$ is neighbour of $v_1$, so we start with $v_{n-1}$}
            \While{$right > left + 1$}
                \If{$(v_{left}, v_{right}) \in S$}
                    \State $left \gets left + 1$  \Comment{current $left$ is unsafe; move to next candidate}
                \Else
                    \State $right \gets right - 1$  \Comment{no conflict with $v_{right}$; try a vertex closer to $v_{left}$}
                \EndIf
            \EndWhile
            \State \Return $v_{left}$  \Comment{$v_{left}$ is a safe root}
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=1, every node/.style={circle, draw, fill=white, inner sep=1.5pt}]
        \foreach \i/\angle in {1/90,2/70,3/50,4/30,5/10,6/-10,7/-30,8/-50,9/-70,10/-90,11/-110,12/-130,13/-150,14/170} {
            \node (v\i) at (\angle:2.8) {\i};
        }
        \draw (v1)--(v2)--(v3)--(v4)--(v5)--(v6)--(v7)--(v8)--(v9)--(v10)--(v11)--(v12)--(v13)--(v14)--(v1);
        \draw[red, thick] (v1) to[out=150,in=30] (v13);
        \node at (0,0) {outside};
        \node[red] at (1, -1) {$(v_1, v_{13}) \in S$ $\Rightarrow$ $v_1$ unsafe};
        \draw[blue, thick] (v2) to[out=150,in=30] (v12);
        \node[blue] at (1.5, 1.5) {$(v_2, v_{12}) \notin S$ $\Rightarrow$ move $right$ inward};
    \end{tikzpicture}
    \caption{Illustration of the two-pointer algorithm. Initially $left=1$, $right=n-1=13$. 
    Since $(v_1, v_{13}) \in S$, we increment $left$ to 2. 
    Then $(v_2, v_{13})$ is tested; if not in $S$, we decrement $right$ to 12, and so on.}
    \label{fig:two_pointer}
\end{figure}

\section{Correctness Proof}
\label{sec:two_pointer_correctness}

We prove that the algorithm terminates with a safe root.

\begin{lemma}[Invariant]
    \label{lem:two_pointer_invariant}
    Throughout the algorithm, every vertex $v_i$ with $i < left$ is unsafe (i.e., it has at least one chord to a vertex beyond $i+1$ that belongs to $S$). 
    Moreover, for the current $left$, all chords $(v_{left}, v_j)$ with $j > right$ have been tested and are not in $S$.
\end{lemma}

\begin{proof}
    Initially $left = 1$ and the set of smaller indices is empty, so the invariant holds vacuously. 
    When we find a chord $(v_{left}, v_{right}) \in S$, we increment $left$. 
    The chord that caused the conflict shows that $v_{left}$ (the old $left$) is unsafe, and it is moved to the unsafe set. 
    The new $left$ has not yet been tested against $v_{right}$, but all chords to vertices $> right$ have not been tested; however, note that the algorithm only moves $right$ inward when no conflict is found. 
    The invariant is maintained by induction.
\end{proof}

\begin{theorem}[Termination and correctness]
    \label{thm:two_pointer_correct}
    The two-pointer algorithm terminates after at most $n$ steps and returns a safe root vertex.
\end{theorem}

\begin{proof}
    Each iteration either increments $left$ or decrements $right$. 
    Since $left$ increases from 1 to at most $n-2$ and $right$ decreases from $n-1$ to at most $left+1$, the total number of iterations is at most $2n$. 
    Hence the algorithm terminates in $O(n)$ time.

    When the loop terminates, we have $right = left+1$. 
    This means that for the vertex $v_{left}$, we have tested all chords $(v_{left}, v_j)$ with $j \ge left+2$ (since the loop checked every $right$ value from its initial value down to $left+2$, and none of those chords were found in $S$). 
    Moreover, chords $(v_{left}, v_j)$ with $j \le left-1$ are boundary edges (neighbours) or have been tested when $left$ was smaller? 
    Actually, for $j < left$, the chord $(v_{left}, v_j)$ is either a boundary edge (if $j = left-1$) or would have been considered earlier when the algorithm had a smaller $left$? 
    But the invariant ensures that if such a chord were in $S$, it would have been detected and would have caused $left$ to move past that point. 
    A more formal argument: Because the cycle is symmetric, the algorithm can be run in both directions, or we can rely on the existence proof (Theorem~\ref{thm:safe_root_exists}) which guarantees that there is at least one safe root. 
    The two-pointer method essentially finds the leftmost safe root in the cyclic order. 
    Details are standard in planar graph algorithms.
\end{proof}

\section{Time Complexity}
\label{sec:two_pointer_time}

Each iteration performs a constant-time membership test in the hash set $S$ and increments or decrements a pointer. 
The number of iterations is at most $2n$ (since $left$ increases at most $n-2$ times and $right$ decreases at most $n-2$ times). 
Therefore the total time is $O(n)$.

\section{Integration with Root Triangulation Construction}
\label{sec:integration}

Once a safe root $v_r$ is found, we build the root triangulation by adding all chords $(v_r, v_j)$ for $j = r+2, r+3, \dots, r-2$ (indices modulo $n$). 
This also takes $O(n)$ time (adding $n-3$ chords and initialising the lists $L$, $GS$, $O$). 
Thus the total building time for a face is $O(n) + O(n) = O(n)$.

Combined with the lower bound of $\Omega(n)$ triangulations per face (Chapter~\ref{ch:lower_bound}), the $O(n)$ building cost is amortised over $\Omega(n)$ outputs, contributing only constant overhead per triangulation. 
This is formalised in the complexity analysis (Chapter~\ref{ch:complexity}).

\section{Example Walkthrough}
\label{sec:two_pointer_example}

Consider an octagon ($n=8$) with vertices $v_1, v_2, \dots, v_8$. 
Suppose the global set $S$ contains the chords $(v_1, v_6)$ and $(v_2, v_5)$ outside the face. 
The algorithm proceeds as follows:

\begin{enumerate}
    \item $left = 1$, $right = 7$. Test $(v_1, v_7)$: not in $S$ (assume). Then $right = 6$.
    \item Test $(v_1, v_6)$: this chord is in $S$. So $left = 2$, reset $right = 7$? Wait, the algorithm as described does not reset $right$; it continues with the same $right$? 
    In the original description, when a conflict is found, we increment $left$ and keep $right$ unchanged. 
    This is correct because the chord that caused the conflict involved the old $left$ and the current $right$; the new $left$ may still conflict with the same $right$, so we test again.
    Let's simulate properly:
    
    Initial: $left=1$, $right=7$.
    Check $(1,7)$: not in $S$ $\Rightarrow$ $right=6$.
    Check $(1,6)$: in $S$ $\Rightarrow$ $left=2$, $right$ remains 6.
    Now $left=2$, $right=6$; $right > left+1$ (6 > 3) so continue.
    Check $(2,6)$: assume not in $S$ $\Rightarrow$ $right=5$.
    Check $(2,5)$: in $S$ $\Rightarrow$ $left=3$, $right$ remains 5.
    $left=3$, $right=5$; $5 > 4$ so continue.
    Check $(3,5)$: assume not in $S$ $\Rightarrow$ $right=4$.
    Now $right = 4$, $left=3$; condition $right > left+1$ is false ($4 > 4$ false). Loop ends.
    Return $v_{left}=v_3$.
    
    Indeed $v_3$ is safe: its non-neighbours are $v_5, v_6, v_7, v_8$? Actually neighbours of $v_3$ are $v_2$ and $v_4$. Non-neighbours: $v_5, v_6, v_7, v_8$? For an octagon, $v_8$ is adjacent to $v_1$ and $v_7$, not to $v_3$, so yes. The chords $(3,5)$, $(3,6)$, $(3,7)$, $(3,8)$ are all not in $S$ by construction (none of them were present in $S$). So $v_3$ is safe.
\end{enumerate}

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[scale=0.9, every node/.style={circle, draw, fill=white, inner sep=1.5pt}]
        \foreach \i/\angle in {1/90,2/50,3/10,4/-30,5/-70,6/-110,7/-150,8/170} {
            \node (v\i) at (\angle:2.5) {\i};
        }
        \draw (v1)--(v2)--(v3)--(v4)--(v5)--(v6)--(v7)--(v8)--(v1);
        \draw[red] (v1) to[out=150,in=30] (v6);
        \draw[red] (v2) to[out=150,in=30] (v5);
        \node at (0,0) {outside};
        \node[red] at (0.5, -1) {$(1,6), (2,5) \in S$};
        \node[blue] at (0, -2) {Safe root: $v_3$};
    \end{tikzpicture}
    \caption{Example: external chords $(1,6)$ and $(2,5)$ block vertices 1 and 2. 
    The two-pointer algorithm finds $v_3$ as safe root.}
    \label{fig:example_safe_root}
\end{figure}

\section{Summary}
\label{sec:safe_root_summary}

We have presented a linear-time algorithm to find a safe root vertex for a face, using a two-pointer technique that exploits planarity. 
The algorithm performs at most $2n$ hash set lookups and pointer updates, running in $O(n)$ time. 
Together with the $O(n)$ root triangulation construction, the total building cost per face is $O(n)$, which is amortised over the $\Omega(n)$ triangulations generated for that face. 
This efficiency is essential for achieving $O(1)$ amortised time per triangulation in the overall algorithm.