% thesis.tex -- Main LaTeX backbone for the thesis
% "Generating All Triangulations of Biconnected Plane Graphs
%  in Amortized Constant Time Per Triangulation"
% Author: Tawkir Aziz Rahman
% Department of Computer Science and Engineering, BUET

\documentclass[12pt, a4paper, twoside]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric, calc, decorations.pathreplacing}
\usepackage{subcaption}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
}
\usepackage{cite}
\usepackage{float}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{setspace}
\onehalfspacing

% Theorem environments
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]
\theoremstyle{remark}
\newtheorem{remark}{Remark}[chapter]

% Algorithm numbering
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

% Title and author
\title{Generating All Triangulations of Biconnected Plane Graphs\\in Amortized Constant Time Per Triangulation}
\author{Tawkir Aziz Rahman\\
Department of Computer Science and Engineering\\
Bangladesh University of Engineering and Technology (BUET)\\
Dhaka-1000, Bangladesh}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We present an algorithm for generating all triangulations of a biconnected plane graph $G$ with $n$ vertices in $O(1)$ amortized time per triangulation and $O(n)$ space. 
Our algorithm establishes a double-layer tree structure consisting of a recursive tree of faces and, for each face, a local genealogical tree of triangulations. 
The key innovation is a novel ``safe root vertex'' selection strategy that prevents the multi-edge problem inherent in biconnected graphs, thereby extending the optimal complexity previously achievable only for triconnected graphs. 
We prove that every face contains at least two safe root vertices regardless of prior triangulations, and we show how to find one in $O(n)$ time using a two‑pointer technique. 
A dynamic global chord set together with a valid generating set (VGS) allows constant‑time edge flips and safe pruning of invalid branches. 
Extensive experiments on 35 diverse test cases validate the theoretical bounds: the algorithm generates tens of millions of triangulations while maintaining nanosecond‑level per‑triangulation time and linear memory usage.
To the best of our knowledge, this is the first algorithm for generating all triangulations of a biconnected plane graph with $O(1)$ amortized time per triangulation.
\end{abstract}

\chapter*{Acknowledgements}
\begin{flushleft}
I express my sincere gratitude to my supervisor for his invaluable guidance and encouragement throughout this research. 
I also thank the authors of the foundational work on triconnected plane graph triangulations, Mohammad Tanvir Parvez, Md. Saidur Rahman, and Shin‑ichi Nakano, whose elegant algorithm provided the starting point for this investigation. 
Finally, I am grateful to the Department of Computer Science and Engineering, BUET, for providing the necessary resources and a conducive research environment.
\end{flushleft}

\tableofcontents
\listoffigures
\listofalgorithms

% ------------------------------
%  Chapters (to be supplied later)
% ------------------------------
\include{chapters/chapter1}   % Introduction
\include{chapters/chapter2}   % Preliminaries
\include{chapters/chapter3}   % Generating All Triangulations of a Cycle (Baseline)
\include{chapters/chapter4}   % Distinction of Our Naming of GS from the Original Paper
\include{chapters/chapter5}   % What the Paper Did and Why It Cannot Be Used for Biconnected Graphs
\include{chapters/chapter6}   % Approaching Faces Differently – Pruning Strategy
\include{chapters/chapter7}   % Multi-Layer DFS and Processing Faces One by One
\include{chapters/chapter8}   % The Problem with Generating Chords and Multi-Edge Conflicts
\include{chapters/chapter9}   % Lower Bound – At Least O(n) Triangulations Per Cycle
\include{chapters/chapter10}  % Finding the Safe Root in O(n) Time (Two-Pointer Algorithm)
\include{chapters/chapter11}  % Maintaining the Valid Generating Set (VGS) and Implementation
\include{chapters/chapter12}  % The Complete Algorithm
\include{chapters/chapter13}  % Complexity Analysis
\include{chapters/chapter14}  % Experimental Validation
\include{chapters/chapter15}  % Why We Cannot Extend This Algorithm to 1-Connected Graphs
\include{chapters/chapter16}  % Future Work
\include{chapters/chapter17}  % Conclusion

% ------------------------------
% Bibliography
% ------------------------------
\bibliographystyle{plain}
\bibliography{references}

\end{document}